Freezing of a Weeks-Chandler-Anderson Fluid using classical Density Functional Theory

### 2020-11-10 papers/thesis-kirstie/thesis.tex
edits and odg file

\section{Soft Fundamental Measure Theory (SFMT)}
In order to apply classical Density Functional Theory to a system that has
a fixed temperature, volume, and nubmer of particles, a functional for 
the Helmholtz free energy must first be formed. An expression for the ideal 
portion of the Helmnoltz free energy is given by Equation~\ref{eq:Fideal-n(r)}, 
but forming an expression for the excess Helmholtz free
energy for the particular fluid being modeled 
from Equation~\ref{eq:Fexcess-simplified} can be challenging. 
Soft Fundamental Measure Theory (SFMT) is a method for forming an 
expression for the excess Helmholtz free energy of a Soft-sphere fluid,
and will be adapted in the Methods section of this paper to fit the 
soft-sphere WCA fluid.
SFMT is an extension of Fundamental Measure Theory (FMT)
that was developed for the Hard-sphere fluid, and which can be derived from
Scaled Particle Theory (SPT).  
%For the Hard Sphere fluid, there is another way to form an expression for 
%the excess Helmholtz free energy in terms of a homogeneous number density 
%and, that is to use Scaled Particle Theory (SPT)~\cite{ReissSPT}.

Scaled Particle Theory seeks a general expression for the excess chemical 
potential of a homogeneous Hard Sphere fluid which can then be integrated 
to find the excess Helmholtz free energy according to the relation
%\begin{equation}F_{ex}=\int{\mu_{ex}dN}\end{equation}
%\begin{equation}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}\end{equation}
\begin{equation}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}{~~~~~}\rightarrow{~~~~~}F_{ex}=\int{\mu_{ex}dN}\end{equation}
It is an extension of the work done by Widom who showed that the excess 
chemical potential of a homogenesous fluid corredsponds to the reversible 
work of inserting a sphere of radius R into a fluid consisting of hard 
spheres of radius R. 
\begin{displaymath}\mu_{ex}=W_{rev}(R){~~~~}\text{work to insert a hard sphere of radius R}\end{displaymath}
Widom's result was obtained by considering the difference 
in the excess Helmholtz free energy when one atom is added to an otherwise 
closed system. He showed that the excess chemcial potential is related to 
the ensemble average of the Boltzman factor for the potential energy 
difference $V_{diff}$ between a system with N atoms and one with N+1 atoms. 
The resulting equation is known as Widom's Insertion Formula.
%\begin{displaymath}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}{~~~~~~}\end{displaymath}
\begin{equation}\mu_{ex}=\frac{F_{ex}(N+1)-F_{ex}(N)}{\Delta{N}}\bigg|_{T,V}\end{equation}
\begin{equation}\label{widoms-insertion-formula}{~}=-k_BT\ln\left(<e^{-\beta{V_{diff}}}>\right)\end{equation}
Considering that at each point in the fluid, there is a probability that 
either a hard sphere of radius R could be inserted, or not, and using 
these two probablities to evaluate $<e^{-\beta{V_{diff}}}>$ this becomes
\begin{equation}{~}=-k_BT\ln\left(e^{-\beta{V_{diff}}}+1\right)\end{equation}
\begin{equation}{~}=V_{diff}=W_{rev}(R)\end{equation}
Returning to Widom's insertion formula Equation~\ref{widoms-insertion-formula},
and considering that it takes no energy for a hard sphere to be placed at a point
in the fluid where it does not overlap with any other spheres, but would require
infinite energy to place a sphere where it would overlap another sphere, 
Equation~\ref{widoms-insertion-formula} becomes
\begin{equation}\mu_{excess}=-k_BT\ln(1-n\frac{4\pi}{3}R_{cavity}^3)\end{equation}
where the probability of a position existing in a homogeneous Hard-sphere fluid 
with number density n that a hard sphere of radius R can be placed without 
overlapping another sphere is given by the number density times the volume 
of the sphere, $nV_{Sphere}$.

%where the radius of the cavity $R_{cavity}$ is equal to the radius of a 
%hard sphere, R, and $<e^{-\beta{V_{diff}}}>$ is evaluated using the 
%probablities that the cavity is either filled or not filled~\cite{Hansen}.
%For cavities that are much larger than one of the hard spheres, 
Forming a more general result for different sizes of cavities in the fluid, 
Scaled Partical Theory arrives at an expression for the excess Helmholtz 
free energy of a homogeneous fluid~\cite{Hansen} in terms of the packing fraction $\eta$ 
\begin{equation}\label{Fexcess-SPT}{\frac{\beta{F_{excess}}}{N}=-ln(1-\eta)+\frac{3\eta}{1-\eta}+\frac{3{\eta}^2}{2(1-\eta)^2}}\end{equation} 
where
\begin{equation}{\eta = \frac{\mbox{Volume occupied by the spheres}}{\mbox{Total Volume}}}\end{equation}
\begin{equation}{\eta = \frac{N\frac{4}{3}\pi{R^3}}{V}}\end{equation}

>>>>

\section{Soft Fundamental Measure Theory (SFMT)}
In order to apply classical Density Functional Theory to a system that has
a fixed temperature, volume, and nubmer of particles, a functional for 
the Helmholtz free energy must first be formed. An expression for the ideal 
portion of the Helmnoltz free energy is given by Equation~\ref{eq:Fideal-n(r)}, 
but forming an expression for the excess Helmholtz free
energy from Equation~\ref{eq:Fexcess-simplified} for the particular fluid 
being modeled can be challenging. 
Soft Fundamental Measure Theory (SFMT) is a method for forming an 
expression for the excess Helmholtz free energy of a Soft-sphere fluid
that does not depend on  explicitly taking into account the interaction 
that each sphere with every other sphere in the fluid as required by 
Equation~\ref{eq:Fexcess-simplified}. In the next chapter on methods, we
will use an SFMT functional that we adapt to fit the soft-sphere WCA fluid.

SFMT is an extension of Fundamental Measure Theory (FMT)
that was developed for the Hard-sphere fluid, and which can be derived from
Scaled Particle Theory (SPT).  
%For the Hard Sphere fluid, there is another way to form an expression for 
%the excess Helmholtz free energy in terms of a homogeneous number density 
%and, that is to use Scaled Particle Theory (SPT)~\cite{ReissSPT}.
Scaled Particle Theory seeks a general expression for the excess chemical 
potential of a homogeneous Hard Sphere fluid which can then be integrated 
to find the excess Helmholtz free energy according to the relation
%\begin{equation}F_{ex}=\int{\mu_{ex}dN}\end{equation}
%\begin{equation}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}\end{equation}
\begin{equation}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}{~~~~~}\rightarrow{~~~~~}F_{ex}=\int{\mu_{ex}dN}\end{equation}
It is an extension of the work done by Widom who showed that the excess 
chemical potential of a homogenesous fluid corredsponds to the reversible 
work of inserting a sphere of radius R into a fluid consisting of hard 
spheres of radius R. 
\begin{align}
	\mu_{ex}=W_{rev}(R){~~~~}\text{work to insert a hard sphere of radius R}
\end{align}
Widom's result was obtained by considering the difference 
in the excess Helmholtz free energy when one atom is added to an otherwise 
closed system. He showed that the excess chemcial potential is related to 
the ensemble average of the Boltzman factor for the potential energy 
difference $V_{diff}$ between a system with N atoms and one with N+1 atoms. 
The resulting equation, shown below, is known as Widom's Insertion 
Formula~\cite{Hansen}.
%\begin{displaymath}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}{~~~~~~}\end{displaymath}
\begin{equation}\mu_{ex}=\frac{F_{ex}(N+1)-F_{ex}(N)}{\Delta{N}}\bigg|_{T,V}\end{equation}
\begin{equation}\label{widoms-insertion-formula}{~}=-k_BT\ln\left(<e^{-\beta{V_{diff}}}>\right)\end{equation}
The quantity $<e^{-\beta{V_{diff}}}>$ can be evaluated by considering
that at each point in the fluid, there is a probability that 
a hard sphere of radius R has space available to exist at that point
without overlapping with other spheres, or it does not, as shown in 
Figure~\ref{fig:p_overlap}.

\begin{figure}[h!]
    \centering
    \includegraphics[height=7cm]{P_overlap.pdf}
    \caption{No energy is needed to place a sphere at a position where 
    there is no overlap such as shown by the black dot at the center of 
    the dotted region near the top. But it is not possible to place a 
    sphere at a position where an overlap occurs.}
    \label{fig:p_overlap}
  \end{figure}

Widom's Insertion formula Equation~\ref{widoms-insertion-formula}, then becomes
\begin{align}
	\mu_{ex} &= -k_BT\ln\left(e^{-\beta{V_{diff}}}+1\right)  \\
	         &= V_{diff} \\
             &= W_{rev}(R)
\end{align}
The quantity $<e^{-\beta{V_{diff}}}>$ can also be evaluated by considering
that it takes no energy for a hard sphere to be placed at a point
in the fluid where it does not overlap with any other spheres 
($V_{diff}=0$), but it would require infinite energy ($V_{diff}=\infty$)
to place a sphere where it would overlap another sphere. 
In this case Equation~\ref{widoms-insertion-formula} becomes
%\begin{equation}\mu_{excess}=-k_BT\ln(1-n\frac{4\pi}{3}R_{cavity}^3) &  \mbox{for} & R_{\cavity}=R\end{equation}
\begin{align}
  \mu_{excess}(R_{cavity}=R)=-k_BT\ln(1-n\frac{4\pi}{3}R^3) 
\end{align}
where the radius of the cavity $R_{cavity}$ equals the radius of a hard sphere R,
and $nV_{Sphere}$ is the probability that a sphere cannot be placed at
some random point in the homogeneous fluid of number density, $n$, without
overlapping with another sphere.

%where the radius of the cavity $R_{cavity}$ is equal to the radius of a 
%hard sphere, R, and $<e^{-\beta{V_{diff}}}>$ is evaluated using the 
%probablities that the cavity is either filled or not filled~\cite{Hansen}.
%For cavities that are much larger than one of the hard spheres, 
Forming a more general result by requiring 
continuity at the boundaries between equations for fluids with 
cavities of a radius $\operatorname{R}_{cavity}$, that is larger, or very 
much larger than the hard sphere radius R, Scaled Partical Theory arrives at an 
expression~\cite{Hansen} for the excess Helmholtz 
free energy of a homogeneous fluid in terms of the packing fraction $\eta$ 
\begin{equation}\label{Fexcess-SPT}{\frac{\beta{F_{excess}}}{N}=-ln(1-\eta)+\frac{3\eta}{1-\eta}+\frac{3{\eta}^2}{2(1-\eta)^2}}\end{equation} 
where
\begin{equation}{\eta = \frac{\mbox{Volume occupied by the spheres}}{\mbox{Total Volume}}}\end{equation}
\begin{equation}{\eta = \frac{N\frac{4}{3}\pi{R^3}}{V}}\end{equation}
